# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

```{r loadData, echo = TRUE}

# Set temp variables to store zipped file and extract directory
tmpzip = paste0 ("tmp", format (Sys.time (), "%Y%m%d%H%M%S"), ".zip")
tmpdir = paste0 ("tmp", format (Sys.time (), "%Y%m%d%H%M%S"))

# set the R_WIN_INTERNET2 to TRUE, for the download.file to work on https
# Download the zipped file and unzip in a temp directory
setInternet2(TRUE)

download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", tmpzip)
tmpcsv = unzip(tmpzip, exdir = tmpdir)

# Load the activity data file
activity = read.csv (tmpcsv, colClasses =
                       c("numeric","character","numeric"))

# Delete the temp directory and files
unlink (tmpzip)
unlink (tmpdir, recursive = TRUE)
```


```{r loadDataAlt, echo = FALSE}

# If the loadData chunk above doesn't work because of operating system or other issues, please follow the steps listed below

# 1. Download the zipped file from the url "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" and unzip in the same directory as the .Rmd file
# 2. Comment out or remove the loadData chunk above
# 3. Uncomment the below line and load the activity data file

#activity = read.csv ("activity.csv", colClasses = c("numeric","character","numeric"))

```

```{r fixDataTypes, echo = TRUE}


# Convert character data in date column to a date data type
activity$date = as.Date (activity$date, format = "%Y-%m-%d")

```

## Histogram of the total number of steps taken each day

```{r histOrig}

# Create summary by total number of steps per date, from the original data with missing values
totStepsNA = with (activity, aggregate ((steps), list (date), sum, na.rm = TRUE))
colnames (totStepsNA) <- c("date","steps")

# Plot histogram using the original data
hist (totStepsNA$steps,
      xlab = "Total number of steps", 
      ylab = "Frequency", main = "Histogram of the total number of steps taken each day",
      col = "red")

```

## What is mean total number of steps taken per day?

```{r meanMedian}

# Calculate the mean and median from the original data with missing values
meanTotStepsNA = mean (totStepsNA$steps, na.rm = TRUE)
medianTotStepsNA = median (totStepsNA$steps, na.rm = TRUE)

```

The mean is `r meanTotStepsNA` and the median is `r medianTotStepsNA`

## What is the average daily activity pattern?

```{r intvPlot}

# Create averages by intervals, from the original data with missing values
avgSteps = with (activity, aggregate ((steps), list (interval), mean, na.rm = TRUE))
colnames (avgSteps) <- c("interval","steps")

# Plot interval vs steps
plot (avgSteps$interval, 
      avgSteps$steps, 
      xlab = "Interval", ylab = "Steps", main = "5-minute interval and average number 
      of steps taken", type = "l", cex = .01, lwd = .01, cex.lab = .85)

# Find the interval with the max steps by ranking the data first and selecting the top row for the first rank
avgSteps$rank = ave (-avgSteps$steps, FUN = function (x) rank (x, ties.method = "min"))
intvMaxSteps = head (subset (avgSteps, rank ==1), 1)

```
The interval with the maximum number of steps is `r intvMaxSteps [1,1]` with `r intvMaxSteps [1,2]` average steps


## Imputing missing values

```{r imputeMissing}

# Create temp dataframe that has only the rows with missing values
activityMissing = activity [!complete.cases (activity),]

# Create a new dataframe where the missing values are filled with the average for that 5-minute interval
activityMissingFixed = merge (x = activityMissing, y = avgSteps, by.x = "interval", by.y = "interval")

# Remove unused columns and rename the steps column to match the activity dataframe
activityMissingFixed$steps.x = NULL
activityMissingFixed$rank = NULL
names(activityMissingFixed)[names(activityMissingFixed) == "steps.y"] = "steps"

# Create tidy dataset by combining the clean rows from the original data with the fixed data
activityTidy = rbind (activity [complete.cases (activity),], activityMissingFixed)

# Create summary by total number of steps per date, from the tidy data
totSteps = with (activityTidy, aggregate ((steps), list (date), sum))
colnames (totSteps) <- c("date","steps")

# Calculate the mean and median from the original data with missing values
meanTotSteps = mean (totSteps$steps)
medianTotSteps = median (totSteps$steps)

# Plot histogram using the tidy data
hist (totSteps$steps,
      xlab = "Total number of steps", 
      ylab = "Frequency", main = "Histogram of the total number of steps taken each day
      after imputing missing values",
      col = "blue")

```

Total number of rows with missing values in the dataset is `r nrow(activityMissing)`

After imputing missing values by filling with the average for that 5-minute interval - 

The new mean is `r meanTotSteps` and the new median is `r medianTotSteps`

The mean has changed by `r meanTotSteps - meanTotStepsNA` and the median has changed by `r medianTotSteps - medianTotStepsNA`

## Are there differences in activity patterns between weekdays and weekends?

```{r cmprWdvWe}

# Add factor variable that captures weekday/weekend
activityTidy$dayType = as.factor(ifelse (weekdays(activityTidy[["date"]]) == "Saturday" | weekdays(activityTidy[["date"]]) == "Sunday", "weekend", "weekday"))

# Create averages by day type and intervals
avgStepsDayType = with (activityTidy, aggregate ((steps), list (dayType, interval), mean))

colnames (avgStepsDayType) <- c("dayType", "interval","steps")

# Load library for lattice plots
library(lattice)

# Panel Plot
xyplot (avgStepsDayType$steps ~ avgStepsDayType$interval | avgStepsDayType$dayType,
      xlab = "Interval", ylab = "Number of steps", main = "Comparing activity patterns during weekdays and weekends", type="l", cex = .01, lwd = .01, cex.lab = .85)
```